### 2017-10-25 Draft2/Appending.Dynamic.tex
More work

\begin{lemma}
\label{lem:CapacitySeq}
	Let $\VarList$ by a dynamic array. Let its \textbf{capacity sequence}, $\VarCapacitySeq$, be the capacities $\VarList$ takes on as $n$ items are appended. Then

	\begin{align*}
	\VarCapacitySeq = 0,\ \VarInitCapacity,\ \VarGrowthFactor\VarInitCapacity,\ \VarGrowthFactor^2\VarInitCapacity,\ \ldots\ \VarGrowthFactor^{\lceil \log_{\VarGrowthFactor} n - \log_{\VarGrowthFactor} \VarInitCapacity \rceil}\VarInitCapacity
	\end{align*}
\end{lemma}

@@

\begin{proof}
	The capacity starts out at zero and becomes $\VarInitCapacity$ once one element is appended. From then on, it can only grow by a factor of $\VarGrowthFactor$ via $\FuncGrow$, so if $\VarGrowthFactor^i\VarInitCapacity$ is the current capacity then $\VarGrowthFactor^{i + 1}\VarInitCapacity$ must be the next capacity. By induction, the rest of the sequence is

	\begin{align*}
	\left\{ \VarGrowthFactor^i\VarInitCapacity \right\}_{i = 1}^k
	\end{align*}

	for some $k$. Since the capacity is only as big as is needs to be, namely greater than or equal to $n$, $k$ is the smallest integer for which $\VarGrowthFactor^k\VarInitCapacity \geq n$. Using this property, I now derive $k$.

	\begin{align*}
	\VarGrowthFactor^k\VarInitCapacity &\geq n\\
	\VarGrowthFactor^k &\geq \frac{n}{\VarInitCapacity}\\
	k &\geq \log_{\VarGrowthFactor} n - \log_{\VarGrowthFactor} \VarInitCapacity\\
	k &\geq \log_{\VarGrowthFactor} n - \log_{\VarGrowthFactor} \VarInitCapacity > k - 1\\
	k &= \big\lceil \log_{\VarGrowthFactor} n - \log_{\VarGrowthFactor} \VarInitCapacity \big\rceil
	\end{align*}

	Then the final term in the sequence is $\VarGrowthFactor^{\lceil \log_{\VarGrowthFactor} n - \log_{\VarGrowthFactor} \VarInitCapacity \rceil}\VarInitCapacity$, completing the proof.
\end{proof}

[make lemma]
Now, consider that to grow by $\VarGrowthFactor$, $\FuncGrow$ must be called. So $\VarGrowthFactor^k\VarInitCapacity$ being present in $S_C$ means $\FuncGrow$ was called at size $\VarGrowthFactor^{k - 1}\VarInitCapacity$ for $k \geq 1$. Then $\FuncGrow$ was called at the sizes

\begin{align*}
S_R = 0,\ \VarInitCapacity,\ \VarGrowthFactor\VarInitCapacity,\ \VarGrowthFactor^2\VarInitCapacity,\ \ldots\ \VarGrowthFactor^{\lceil \log_{\VarGrowthFactor} n - \log_{\VarGrowthFactor} \VarInitCapacity \rceil - 1}\VarInitCapacity
\end{align*}

Each time $\FuncGrow$ is called, $i$ items are copied where $i$ is the current size of the dynamic array. Then the total number of items copied for $n$ appends is

\begin{align*}
0 + \VarInitCapacity + \VarGrowthFactor\VarInitCapacity + \ldots + \VarGrowthFactor^{\lceil \log_{\VarGrowthFactor} n - \log_{\VarGrowthFactor} \VarInitCapacity \rceil - 1}\VarInitCapacity &= \left( \frac{\VarGrowthFactor^{\lceil \log_{\VarGrowthFactor} n - \log_{\VarGrowthFactor} \VarInitCapacity \rceil} - 1}{\VarGrowthFactor - 1} \right) \VarInitCapacity
\end{align*}

Finally, counting the writes made for every item in $\FuncAppend$, I finally find an explicit formula for $\FuncWrites(n)$.

[ref1]
\begin{align*}
\FuncWrites(n) = n + \left( \frac{\VarGrowthFactor^{\lceil \log_{\VarGrowthFactor} n - \log_{\VarGrowthFactor} \VarInitCapacity \rceil} - 1}{\VarGrowthFactor - 1} \right) \VarInitCapacity
\end{align*}

[ref2]
To make this expression easier to work with, I will write it in big-P notation to approximate it for large $n$. First I note that $\log_{\VarGrowthFactor} k \leq \lceil \log_{\VarGrowthFactor} k \rceil < \left( \log_{\VarGrowthFactor} k \right) + 1$ for any $k > 0$, so $k \leq \VarGrowthFactor^{\lceil \log_{\VarGrowthFactor} k \rceil} < \VarGrowthFactor{k}$. Then

\begin{align*}
\FuncWrites(n) &= n + \left( \frac{\VarGrowthFactor^{\big\lceil \log_{\VarGrowthFactor} \left( \frac {n} {\VarInitCapacity} \right) \big\rceil} - 1}{\VarGrowthFactor - 1} \right) \VarInitCapacity\\
n + \left( \frac{\frac{n}{\VarInitCapacity} - 1}{\VarGrowthFactor - 1} \right) \VarInitCapacity \leq \FuncWrites(n) &< n + \left( \frac{\VarGrowthFactor\frac{n}{\VarInitCapacity} - 1}{\VarGrowthFactor - 1} \right) \VarInitCapacity\\
n + \left( \frac{n - \VarInitCapacity}{\VarGrowthFactor - 1} \right) \leq \FuncWrites(n) &< n + \left( \frac{\VarGrowthFactor{n} - \VarInitCapacity}{\VarGrowthFactor - 1} \right)\\
P\left( n + \left( \frac{n - \VarInitCapacity}{\VarGrowthFactor - 1} \right) \right) \leq P\left( \FuncWrites(n) \right) &< P\left( n + \left( \frac{\VarGrowthFactor{n} - \VarInitCapacity}{\VarGrowthFactor - 1} \right) \right)\\
P\left( \left( \frac{\VarGrowthFactor}{\VarGrowthFactor - 1} \right) n \right) \leq P\left( \FuncWrites(n) \right) &< P\left( \left( \frac{2\VarGrowthFactor - 1}{\VarGrowthFactor - 1} \right) n \right)
\end{align*}

@@

I wish to find the space allocated while appending $n$ items to a dynamic array. I denote this quantity $\FuncSpace(n)$, and define it as the total length of all arrays allocated during $n$ $\FuncAppend$ calls. I now derive a formula for $\FuncSpace(n)$.

From \ref{lem:CapacitySeq} I know the exact capacities $\VarList$ takes on as $n$ items are appended. Dynamic arrays also have the property that $i \in S_C$ iff an array of length $i$ was allocated. Then the total length of arrays allocated is

\begin{align*}
\FuncSpace(n) &= \sum_{i \in S_C} i\\
&= 0 + \VarInitCapacity + \VarGrowthFactor\VarInitCapacity + \VarGrowthFactor^2\VarInitCapacity + \ldots + \VarGrowthFactor^{\lceil \log_{\VarGrowthFactor} n - \log_{\VarGrowthFactor} \VarInitCapacity \rceil}\VarInitCapacity\\
&= \left( \frac{\VarGrowthFactor^{\lceil \log_{\VarGrowthFactor} n - \log_{\VarGrowthFactor} \VarInitCapacity \rceil + 1} - 1}{\VarGrowthFactor - 1} \right) \VarInitCapacity
\end{align*}

Using the identity in [ref2] again, I bound its big-P complexity:

\begin{align*}
\left( \frac{\VarGrowthFactor \frac{n}{\VarInitCapacity} - 1}{\VarGrowthFactor - 1} \right) \VarInitCapacity \leq \FuncSpace(n) &< \left( \frac{\VarGrowthFactor^2 \frac{n}{\VarInitCapacity} - 1}{\VarGrowthFactor - 1} \right) \VarInitCapacity\\
P\left( \left( \frac{\VarGrowthFactor \frac{n}{\VarInitCapacity} - 1}{\VarGrowthFactor - 1} \right) \VarInitCapacity \right) \leq P\left( \FuncSpace(n) \right) &< P\left( \left( \frac{\VarGrowthFactor^2 \frac{n}{\VarInitCapacity} - 1}{\VarGrowthFactor - 1} \right) \VarInitCapacity \right)\\
P\left( \left( \frac{\VarGrowthFactor}{\VarGrowthFactor - 1} \right) n \right) \leq P\left( \FuncSpace(n) \right) &< P\left( \left( \frac{\VarGrowthFactor^2}{\VarGrowthFactor - 1} \right) n \right)
\end{align*}

>>>>

\begin{lemma}
\label{lem:CapacitySeq}
	Let $\VarList$ by a dynamic array. Let its \textbf{capacity sequence}, $\VarCapacitySeq$, be the capacities $\VarList$ takes on as $n$ items are appended. Then
	\begin{align*}
	\VarCapacitySeq = 0,\ \VarInitCapacity,\ \VarGrowthFactor\VarInitCapacity,\ \VarGrowthFactor^2\VarInitCapacity,\ \ldots\ \VarGrowthFactor^{\lceil \log_{\VarGrowthFactor} n - \log_{\VarGrowthFactor} \VarInitCapacity \rceil}\VarInitCapacity
	\end{align*}
\end{lemma}

@@

\begin{proof}
	The capacity starts out at zero and becomes $\VarInitCapacity$ once one element is appended. From then on, it can only grow by a factor of $\VarGrowthFactor$ via $\FuncGrow$, so if $\VarGrowthFactor^i\VarInitCapacity$ is the current capacity then $\VarGrowthFactor^{i + 1}\VarInitCapacity$ must be the next capacity. By induction, the rest of the sequence is $\left( \VarGrowthFactor^i\VarInitCapacity \right)_{i = 1}^\VarUseful$ for some $\VarUseful$.

	Since the capacity is only as big as is needs to be, namely greater than or equal to $n$, $\VarUseful$ is the smallest integer for which $\VarGrowthFactor^\VarUseful\VarInitCapacity \geq n$. Using this property, I now derive $\VarUseful$.
	\begin{align*}
	\VarGrowthFactor^\VarUseful\VarInitCapacity &\geq n > \VarGrowthFactor^{\VarUseful - 1}\VarInitCapacity\\
	\VarGrowthFactor^\VarUseful &\geq \frac{n}{\VarInitCapacity} > \VarGrowthFactor^{\VarUseful - 1}\\
	\VarUseful &\geq \log_{\VarGrowthFactor} n - \log_{\VarGrowthFactor} \VarInitCapacity > \VarUseful - 1\\
	\end{align*}
	Since $\VarUseful$ is an integer,
	\begin{align*}
	\VarUseful &= \big\lceil \log_{\VarGrowthFactor} n - \log_{\VarGrowthFactor} \VarInitCapacity \big\rceil
	\end{align*}
	The final term in the sequence is $\VarGrowthFactor^\VarUseful = \VarGrowthFactor^{\lceil \log_{\VarGrowthFactor} n - \log_{\VarGrowthFactor} \VarInitCapacity \rceil}\VarInitCapacity$, completing the proof.
\end{proof}

[make lemma]
Now, consider that to grow by $\VarGrowthFactor$, $\FuncGrow$ must be called. So $\VarGrowthFactor^i\VarInitCapacity$ being present in $S_C$ means $\FuncGrow$ was called at size $\VarGrowthFactor^{i - 1}\VarInitCapacity$ for $i \geq 1$. Then $\FuncGrow$ was called at the sizes
\begin{align*}
\VarGrowSeq = 0,\ \VarInitCapacity,\ \VarGrowthFactor\VarInitCapacity,\ \VarGrowthFactor^2\VarInitCapacity,\ \ldots\ \VarGrowthFactor^{\VarUseful - 1}\VarInitCapacity
\end{align*}
Each time $\FuncGrow$ is called, $s$ items are copied where $s$ is the current size of the dynamic array. Then the total number of items copied for $n$ appends is
\begin{align*}
\sum_{s \in \VarGrowSeq} s &= 0 + \VarInitCapacity + \VarGrowthFactor\VarInitCapacity + \ldots + \VarGrowthFactor^{\VarUseful - 1}\VarInitCapacity\\
&= \left( \frac{\VarGrowthFactor^{\VarUseful} - 1}{\VarGrowthFactor - 1} \right) \VarInitCapacity
\end{align*}
Counting the writes made for every item in $\FuncAppend$, I finally find an explicit formula for $\FuncWrites(n)$.
[ref1]
\begin{align*}
\FuncWrites(n) = n + \left( \frac{\VarGrowthFactor^\VarUseful - 1}{\VarGrowthFactor - 1} \right) \VarInitCapacity
\end{align*}
This expression looks clean at a glance, but $\VarUseful$ implicitly depends on $n$ and uses the ceiling function and logarithms. To make it easier to analyze, I will asymptotically bound this expression.

[ref2]
First I note that $\log_{\VarGrowthFactor} k \leq \lceil \log_{\VarGrowthFactor} k \rceil < \left( \log_{\VarGrowthFactor} k \right) + 1$ for any $k > 0$, so $k \leq \VarGrowthFactor^{\lceil \log_{\VarGrowthFactor} k \rceil} < \VarGrowthFactor{k}$. Since $\VarUseful = \left\lceil \log_{\VarGrowthFactor} \frac{n}{\VarInitCapacity} \right\rceil$, $\frac n \VarInitCapacity \leq \VarGrowthFactor^\VarUseful < \frac {\VarGrowthFactor{n}} \VarInitCapacity$. Then
\begin{align*}
\FuncWrites(n) &= n + \left( \frac{\VarGrowthFactor^\VarUseful - 1}{\VarGrowthFactor - 1} \right) \VarInitCapacity\\
n + \left( \frac{n / \VarInitCapacity - 1}{\VarGrowthFactor - 1} \right) \VarInitCapacity \leq \FuncWrites(n) &< n + \left( \frac{\VarGrowthFactor{n} / \VarInitCapacity - 1}{\VarGrowthFactor - 1} \right) \VarInitCapacity\\
n + \left( \frac{n - \VarInitCapacity}{\VarGrowthFactor - 1} \right) \leq \FuncWrites(n) &< n + \left( \frac{\VarGrowthFactor{n} - \VarInitCapacity}{\VarGrowthFactor - 1} \right)\\
n + \left( \frac{n - \VarInitCapacity}{\VarGrowthFactor - 1} \right) \RelLessEqual \FuncWrites(n) &\RelLess n + \left( \frac{\VarGrowthFactor{n} - \VarInitCapacity}{\VarGrowthFactor - 1} \right)\\
\left( \frac{\VarGrowthFactor}{\VarGrowthFactor - 1} \right) n \RelLessEqual \FuncWrites(n) &\RelLess \left( \frac{2\VarGrowthFactor - 1}{\VarGrowthFactor - 1} \right) n
\end{align*}

@@

I wish to find the space allocated when $n$ items are appended to a dynamic array. I call this quantity the \textbf{space cost}, denote it $\FuncSpace(n)$, and define it as the total length of arrays allocated by $n$ $\FuncAppend$ calls. I now derive a formula for $\FuncSpace(n)$.

From \ref{lem:CapacitySeq}, I know the capacity sequence of $\VarList$. Dynamic arrays also have the property that $c \in \VarCapacitySeq$ iff an array of length $c$ was allocated. Then the total length of arrays allocated is

\begin{align*}
\FuncSpace(n) &= \sum_{c \in \VarCapacitySeq} c\\
&= 0 + \VarInitCapacity + \VarGrowthFactor\VarInitCapacity + \VarGrowthFactor^2\VarInitCapacity + \ldots + \VarGrowthFactor^\VarUseful\VarInitCapacity\\
&= \left( \frac{\VarGrowthFactor^{\VarUseful + 1} - 1}{\VarGrowthFactor - 1} \right) \VarInitCapacity
\end{align*}

Using the identity in [ref2] again, I asymptotically bound $\FuncSpace(n)$:

\begin{align*}
\left( \frac{\VarGrowthFactor{n} / \VarInitCapacity - 1}{\VarGrowthFactor - 1} \right) \VarInitCapacity \leq \FuncSpace(n) &< \left( \frac{\VarGrowthFactor^2{n} / \VarInitCapacity - 1}{\VarGrowthFactor - 1} \right) \VarInitCapacity\\
\left( \frac{\VarGrowthFactor{n} / \VarInitCapacity - 1}{\VarGrowthFactor - 1} \right) \VarInitCapacity \RelLessEqual \FuncSpace(n) &\RelLess \left( \frac{\VarGrowthFactor^2{n} / \VarInitCapacity - 1}{\VarGrowthFactor - 1} \right) \VarInitCapacity\\
\left( \frac{\VarGrowthFactor}{\VarGrowthFactor - 1} \right) n \RelLessEqual \FuncSpace(n) &\RelLess \left( \frac{\VarGrowthFactor^2}{\VarGrowthFactor - 1} \right) n
\end{align*}
